---
title: "Diff-in-Diff 1: Simple DiD"
author: "Matt Birch"
date: "2023-02-25"
output: html_document
---

# Intro to Difference in Differences
### This one will be the simple version. More exciting videos to come!

```{r, message = FALSE, warning = FALSE}
rm(list = ls())
library(ggplot2)
library(stargazer)
```

The best approach to estimating causal impacts is with a randomized trial, such as with an A/B test. In many cases, however, randomization is either unfeasible or unethical, or both. Sometimes the real world conducts natural experiments, in which some people receive treatment while others do not. With careful analysis, we can exploit these natural experiments to identify the causal relationship. 

One of the most common methods is Difference in Differences, also called diff-in-diff or DiD. DiD is an empirical technique to exploit natural experiments to identify causal relationships. The general idea is that the data is separated into treatment and control groups, without us running a real experiment. Under certain assumptions, we can compare the treatment and control groups (one difference) before and after treatment occurs (another difference). We can see how the difference between groups changes, and find our causal effect. 

So let's get some [fake] data!

```{r}
stem_data <- read.csv("https://raw.githubusercontent.com/MattBirch42/Causal-Inference-Stuff/main/Fake_DD_stemPolicy_scores.csv")
```
### Explanation of Data
This is fake data about an imagined scenario that I used for a class assignment. The STEM program gives students in the participating region extra training in math and science. Not all regions participate. We will compare the test scores of students in participating regions those who are not before and after the program is implemented. 

Features:

* year = Year
* male = 1 if male, 0 otherwise
* Region: These represent some fictional geographic region
* Race: These categorize different unspecified races.
* Stem: stem denotes extra STEM prep in grades 10-12. Stem program starts in 2013 in region 2 and in 2015 in region 5. No other region has the program.
* bday_m_cutoff: regions 1-4 have compulsory schooling laws (save this for RD later)
* Education: parent’s years of formal schooling
* Income: family income
* Score: Assessment exam score (this is our dependent variable)
* Varsity: did varsity sports
* single_parent_hh: Lives with single parent

I am going to ignore region 5 in this video, since it receives treatment ad a different time than region 2. We'll come back to region 5 in DiD part 3.That means that our treatment group in this video is region 2, and the control is everything else except region 5. 

```{r}
stem2 = stem_data[stem_data$region != 5,]
```

### Create plot to show the issue
Let's look at average scores in region 2 vs other regions combined.

Before 2013, Region 2 seems to lag behind the rest. Then in 2013, with the STEM program, scores in Region 2 jump to a higher level. With a good DiD, you can *see* the treatment in a graph like this. Although it may not be *this* obvious.

However, we still have to talk about assumptions and estimation. 

```{r, echo = FALSE}
stem2$reg2 = ifelse(stem2$region == 2,'Region 2','Other Regions')

avg_scores <- aggregate(score ~ reg2 + year, data = stem2, FUN = mean)

ggplot(avg_scores, aes(x = year, y = score, color = reg2)) +
  geom_point(size = 3) +
  labs(x = "Year", y = "Average Score", color = "Region") +
  theme_light()+
  geom_vline(xintercept = 2012.5, linetype = "dashed") +
  scale_x_continuous(breaks = seq(2010,2019, by = 1))

```

### Assumption
To successfully execute DiD and trust the results as a causal effect, you have to assume that the trends between treatment and control are parallel before and after the treatment. In this case, the trends do seem similar, before and after. 

### Execution
To perform the most basic version of a DiD, you use a regression with at least 3 variables. You must have:

* a treatment group variable that determines whether a group is ever treated
* a treatment time variable that says whether treatment is occurring anywhere (doesn't actually have to be time)
* an interaction term that marks whether a sample is in the treatment group AND ALSO currently being treated.

You can add additional covariates, but these 3 are teh required components.

Treatment group:
``` {r}
stem2$r2<-ifelse(stem2$region==2,1,0)
```

Treatment time:
``` {r}
stem2$y2013plus<-ifelse(stem2$year>=2013,1,0)
```

Interaction, both group and time:
``` {r}
stem2$r2_2013<-stem2$r2*stem2$y2013plus
```

And the actual regression: 
``` {r}
dd_reg1 <- lm(score ~ r2_2013 + r2 + y2013plus, data = stem2)
dd_reg1$coefficients
```

We interpret these coefficients separately, but only the coefficient on the interaction is the actual causal effect from the STEM treatment. 

The intercept tells us that the baseline score in this data set (the control regions) is `r round(dd_reg1$coefficients[1],2)`.

The time `r2` coefficient tells us that in the beginning, Region 2 had scores `r abs(round(dd_reg1$coefficients[3],))` points `r ifelse(dd_reg1$coefficients[3]<0,'below','above')` the baseline of the other regions.

The coefficient on the time variable `y2013plus` says that across regions, `score` was `r round(dd_reg1$coefficients[4],2)` points higher after 2013 than before. 

We have accounted for regions and time trends. Now we use the interaction to see if there is a difference in the regional differences before and after the time change. The coefficient on `r2_2013`, `r round(dd_reg1$coefficients[2],2)` tells us that students in Region 2 increased their scores by `r round(dd_reg1$coefficients[2],2)` points more than the other groups. The difference between groups changed by that much. This would be the causal effect of the STEM program. This is how many points the treatment group gains above and beyond the shared time trend. 

It may be helpful to see it in a graph.


```{r, echo = FALSE}

ggplot(avg_scores, aes(x = year, y = score, color = reg2)) +
  geom_point(size = 3) +
  labs(x = "Year", y = "Average Score", color = "Region") +
  theme_light()+
  geom_vline(xintercept = 2012.5, linetype = "dashed") +
  scale_x_continuous(breaks = seq(2010,2019, by = 1)) + 
  # this is wrong. change time to time trend. 
  geom_segment (aes (x=2010,xend=2012,y=dd_reg1$coefficients[1],yend=dd_reg1$coefficients[1]+dd_reg1$coefficients[4]))

```


#plot with regression lines added on for effect.

# with additional covariates

# stargazer for easy comparison

# inference
#inference
